Typing a linear \picalc

### 2020-04-06 paper/paper.tex
Think about substitution
Solve end case in subject reduction
Add a bit more to the introduction

\newcommand{\stacked}[1]{\mprset{flushleft} \inferrule*{}{#1}}
\newcommand{\datatype}[2]{{\mprset{fraction={===}} \inferrule{#1}{#2}}}

\newcommand{\lamdacalc}{$\lambda$-calculus}
\newcommand{\picalc}{$\pi$-calculus}
\newcommand{\Picalc}{$\pi$-Calculus}

@@

We live in a concurrent world where any given state is often decided by the interaction between a myriad of parties --- people, machines, processors, services, networks.
As humans, we aim to model, predict, and even construct such interactive systems.
As mathematicians, abstraction is our tool of choice to do so.

The \picalc models communication.

The properties guaranteed by the picalc

@@

Machines prefer things they can algorithmically act upon. Like natural numbers!
What does it mean to use a number as a variable reference?
The idea de Bruijn had \cite{} was to use the index $n$ to refer to the variable introduced $n$ binders ago.
The binders themselves introduce no names anymore.
The expression $\lambda g . (\lambda f . f g) g$ in the \lamdacalc{} would translate as $\lambda (\lambda 0 1) 0$.
That is, terms at different \emph{depths} must use different indices to refer to the same binding.
Humans find this often confusing.
There is however no reason not to keep the original names bestowed by the humans together with the indices.
Machines can then manipulate references mechanically and still use names to present them to humans.

@@

In the \lamdacalc{} $\beta$-reduction operates on syntactically adjacent terms.
In the \picalc{} however, the syntax introduces unnecessary distinctions
(e.g. semantically parallel composition is defined modulo associativity and commutativity).
There are several ways around this, a structural congruence relation being one of the historical ones.
(Others include labeled transition systems and higher inductive types.)

@@

A type system is in the business of further constraining a language, filtering out terms that (although well-formed) are meaningless.
In the simply typed \lamdacalc{}, a term has a meaning iff it has a type.
This typing judgment is a proof of the form $\Gamma \vdash t : T$, where the term $t$ has type $T$ under the typing context $\Gamma$ -- the typing context associates variables (possibly occurring within $t$) to types.
In the simply typed \picalc{} however, it is the channels that are typed, and a process has a meaning iff it \emph{uses} its channels according to their types.
The typing judgment is therefore of the form $\Gamma \vdash p$, where the process $p$ (which has no type of its own) uses the channels in $\Gamma$ according to their types.

@@

@@

\todo{syntax directed typing rules}

>>>>

% Abbreviations
\newcommand{\lambdacalc}{$\lambda$-calculus}
\newcommand{\picalc}{$\pi$-calculus}
\newcommand{\Picalc}{$\pi$-Calculus}

% Typing rules
\newcommand{\stacked}[1]{\mprset{flushleft} \inferrule*{}{#1}}
\newcommand{\datatype}[2]{{\mprset{fraction={===}} \inferrule{#1}{#2}}}

@@

We live in a concurrent world where any given state is often computed interactively by a myriad of parties --- people, machines, processors, services, networks.
As humans, we aim to model, predict, and even construct such interactive systems; as mathematicians, abstraction is our tool of choice to do so.
The \picalc{} is a formal language that abstracts over the details of concurrent processing and boils the interactions down to the sending and receiving of data through communication channels.
Notably, it features \emph{channel mobility}: the channels themselves are first order constructs, and can therefore be transmitted too.
We define the syntax for the \picalc{} in \S \ref{syntax}.
We use dependent types to make this syntax well-scoped by construction: every unbound variable is accounted for in the type of the process using it.
\todo{de Bruijn}

In the \lambdacalc{} all computations are local and happen on syntactically contiguous terms; in the \picalc{} however computations need not to happen on syntactically contiguous terms.
One way to define a reduction relation modulo these syntactic minutiae is to define a structural congruence relation that acts as a quotient type.
A term $A$ can then be freely rewritten into a term $B$ as long as $A$ is structurally congruent to $B$.
We introduce this equivalence relation in \S \ref{structural-congruence}.

In the \lambdacalc{} function application is the main computational engine; in the \picalc{} it is inter-process communication what drives computation.
Two subprocesses can communicate anywhere in a process but under input and output, which must be respected to ensure the sequentiality in behavior.
We provide an operational semantics for the \picalc{} in \S \ref{operational-semantics}.
Every computational step is the result of communication; we take care to keep information on what channel the communication is happening on so that we can later use it to state the theorem for subject reduction (aka type preservation).

As with the \lambdacalc{}, the \picalc{} can be used to construct a wide variety of misbehaved processes.

The properties guaranteed by the linear picalc

@@

Machines prefer things they can algorithmically act upon. Like natural numbers!
What does it mean to use a number as a variable reference?
The idea de Bruijn had \cite{} was to use the index $n$ to refer to the variable introduced $n$ binders ago.
The binders themselves introduce no names anymore.
The expression $\lambda g . (\lambda f . f g) g$ in the \lambdacalc{} would translate as $\lambda (\lambda 0 1) 0$.
That is, terms at different \emph{depths} must use different indices to refer to the same binding.
Humans find this often confusing.
There is however no reason not to keep the original names bestowed by the humans together with the indices.
Machines can then manipulate references mechanically and still use names to present them to humans.

@@

In the \lambdacalc{} $\beta$-reduction operates on syntactically adjacent terms.
In the \picalc{} however, the syntax introduces unnecessary distinctions
(e.g. semantically parallel composition is defined modulo associativity and commutativity).
There are several ways around this, a structural congruence relation being one of the historical ones.
(Others include labeled transition systems and higher inductive types.)

@@

A type system is in the business of further constraining a language, filtering out terms that (although well-formed) are meaningless.
In the simply typed \lambdacalc{}, a term has a meaning iff it has a type.
This typing judgment is a proof of the form $\Gamma \vdash t : T$, where the term $t$ has type $T$ under the typing context $\Gamma$ -- the typing context associates variables (possibly occurring within $t$) to types.
In the simply typed \picalc{} however, it is the channels that are typed, and a process has a meaning iff it \emph{uses} its channels according to their types.
The typing judgment is therefore of the form $\Gamma \vdash p$, where the process $p$ (which has no type of its own) uses the channels in $\Gamma$ according to their types.

@@

\todo{add on 0 = y * z -> 0 = y or 0 = z}

@@

\todo{stress what leftover typing is as a technique}
\todo{syntax directed typing rules}
